Renewal and stability in populations structured by remaining years of life

### 2015-10-14 ThanoRepro/ThanoRepro.tex
further massaging of intro

\onehalfspacing
All demographic phenomena vary over age in regular ways. Information
on such phenomena, along with the size and age structure of a population, allows
the demographer to make predictions about the future size and structure of a population. In this
paper, I explore some formal demographic consequences of a particular
redefinition of age. Instead of counting age as the time passed since birth, consider age as the amount of time left
until death\footnote{This quantity is referred to as \textit{residual} lifetime
in the reliability literaure.}. The first kind of age is called
chronological age and the kind of age explored in this paper is
called thanatological age. Individuals in this case move in the same direction
along imaginary life lines, but the reference point for age is now at the end of
the life line instead of at the beginning. For individuals, the timing of events
in life is identified exactly using either of the two definitions of age, but
aggregate patterns differ. Some demographic phenomena are best
described as a function of time since birth, and others of time until death,
while still others can be a function of both to some degree \citep{riffe2015ttd}. This observation ought to motivate the formal demographic base by incorporating this aspect of time in into
models.

%Temporal perspectives and rescalings of the lifecourse are abundant in
% demography already. Such work has enriched the discipline by producing new insights, questions, and conundrums.

Direct information on the timing of demographic transitions with respect to
death is available retrospectively from linked population registers or
panel surveys with mortality follow-ups.
Some surveys also ask respondents in various ways about their perceived future lifetime, which may underly those
demographic transitions governed to some degree by volition. The present inquiry
draws only from lifetable-projected remaining lifetime, which
is intended to approximate actual remaining lifetime under the strong
assumption of homogenous populations.
The present inquiry is based on lifetable methods, which redistribute
chronological age-classified counts over future death times according to a fixed
mortality schedule. This allows us to make statements about the remaining lifelines within a stock of population, under certain assumptions.

% Fertility and reproduction are plausibly a function of both age perspectives,
 % and the question of how to best partition them over time is not investigated here.

In this paper I describe some formal consequences
of the Lotka-Euler renewal model when structured by thanatological age. The
renewal model relates fertility, mortality, and population structure in a parsimonious form, revealing the long term consequences of a particular set of conditions. All diverse lines
of contemporary demographic research are relatable to the renewal model.
Some might think that stable population theory is not relevant to everyday
demography, since populations are always changing. Even if change undercuts the
notion of stability as a prediction, models of long-run consequnces still give a
useful index of the demographic force implied by a given set of rates.
That population projections lead into a stable state means that some
form of stability is implied by the routine practice of demography.
Stability or even stationarity are imaginable for the world population of humans
one day. Models of stationary populations are easily applied by analogy to
closed or nearly-closed birth cohorts. To present the renewal model based on thanatological
age not necessarily a suggestion for a better way to model or project
population. Rather, the thanatological renewal model represents an alternative
and coherent way to describe the process of population renewal. A model relating
renewal to a death flow rather than to a birth flow may have applications, but
I do not speculate on these here.

We have stated that many lifecourse transitions covary more with remaining
lifetime than with life lived. In an even more general sense, by thinking about remaining lifetime as a
particular class of time-to-transition, one may generalize the lifetable methods presented here into new area applications, such as education, partnering, the economic lifecycle, or migration.

The chronological-age classified Lotka-Euler renewal model and the
thanatological renewal model both approximate the movement of population stocks
over time, but from opposite vantage points. Specifying a model of
thanatological renewal expands the toolbox available to approach complex
temporal population structure. The next section of this article describes the lifetable method of transforming age-classified
data into thanatological age classes. I then derive the continuous version of
the thanatological renewal model, and some other stable properties. This is
followed by the matrix version of the model, which is explained alongside the
lifecycle graph. I present a set of preliminary empirical findings to help
understand the implications of temporal structure.

\section*{A lifetable approximation of thanatological age}

>>>>

\onehalfspacing
In this paper, I explore some formal demographic consequences of a particular redefinition of age. Instead of counting age as the time passed since birth, consider age as the amount of time left until death\footnote{This quantity is referred to as \textit{residual} lifetime
in the reliability literaure, or just time-to-death (TTD) elsewhere.}. The first
kind of age is called chronological age and the kind of age explored in this paper is
called thanatological age. Individuals in this case move in the same direction
along imaginary life lines, but the reference point for age is now at the end of
the life line instead of at the beginning. For individuals, the timing of events
in life is identified exactly using either of the two definitions of age, but
aggregate patterns differ. The models of demography are prefaced by the observation
that demographic phenomena vary over age and time in empirically regular ways.
Some such phenomena are best described in the aggregatate as a function
of time since birth, and others of time until death, while still others can be a function of both to some degree \citep{riffe2015ttd}. This observation ought to motivate the formal demographic base by incorporating this aspect of time in into
models.

%Temporal perspectives and rescalings of the lifecourse are abundant in
% demography already. Such work has enriched the discipline by producing new insights, questions, and conundrums.
% Fertility and reproduction are plausibly a function of both age perspectives,
 % and the question of how to best partition them over time is not investigated here.

I describe the Lotka-Euler-Leslie renewal model when structured by
thanatological age. The renewal model relates fertility, mortality, and population structure in a parsimonious form, revealing the long term consequences of a particular set of conditions. All diverse lines
of contemporary demographic research are relatable to the renewal model.
Some might say that stable population theory is not relevant to everyday
demography, since populations are always changing. Even if change precludes the
notion of stability as a prediction, models of long-run consequnces still give a
useful measure of the demographic force implied by a given set of rates.
In this and other ways, stability is never far away. That population projections
lead into a stable state means that some form of stability is implied by the routine practice of demography.
Stability or even stationarity are imaginable for the world population of humans
one day. Models of stationary populations are easily applied by analogy to
closed or nearly-closed birth cohorts. Finally, stable population theory is
relavant to other non-human structured populations, such as stars.

%We have stated that many lifecourse transitions covary more with remaining
%lifetime than with life lived. In an even more general sense, by thinking about
% remaining lifetime as a particular class of time-to-transition, one may generalize the lifetable methods presented here into new area applications, such as education, partnering, the economic lifecycle, or migration.

To present the renewal model based on thanatological age is not necessarily a
suggestion for a better way to model or project population. Rather, the
thanatological renewal model represents an alternative and coherent way to
describe the process of population renewal by relating renewal to a death flow
rather than to a birth flow. The chronological-age classified renewal model and the thanatological renewal model both approximate the movement of population stocks
over time, but from opposite vantage points. Specifying a model of
thanatological renewal expands the toolbox available to approach complex
temporal population structure. The next section of this article describes the lifetable method of transforming age-classified
data into thanatological age classes. I then derive the continuous version of
the thanatological renewal model, and some other stable properties. This is
followed by the matrix version of the model, which is explained alongside the
lifecycle graph.

\section*{A lifetable approximation of thanatological age}
Direct information on the timing of demographic transitions with respect to
death is available retrospectively from linked population registers or
panel surveys with mortality follow-ups.
Some surveys also ask respondents in various ways about their perceived future lifetime, which may underly those
demographic transitions governed to some degree by volition. The present inquiry
draws only from lifetable-projected remaining lifetime, which
is intended to approximate actual remaining lifetime under the strong
assumption of homogenous populations. Chronological age-classified counts are
redistributed over future death times according to a fixed mortality schedule.
This allows us to make statements about the remaining lifetime distribution
within a stock of population, under certain assumptions.
